---
title: "SeylerHyyppa_FinalProject_datawrangling"
author: "Reino Hyyppa & Meghan Seyler"
date: "3/31/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---
Reminders for knitting commands in code chunks - can delete when we're done

* Knitting commands in code chunks:
 - `include = FALSE` - code is run, but neither code nor results appear in knitted file
 - `echo = FALSE`    - code not included in knitted file, but results are
 - `eval = FALSE`    - code is not run in the knitted file
 - `message = FALSE` - messages do not appear in knitted file
 - `warning = FALSE` - warnings do not appear...
 - `fig.cap = "..."` - adds a caption to graphical results

# 1. Introduction 

## 1.1 Context 

Waste diversion comes with global benefits including a decrease in  
global warming potential (GWP). However, waste diversion stands to especially 
benefit those populations living closest to landfills by decreasing local 
green house gas emissions and other hazards such as odor, smoke, noise, and  
water contamination.This project aims to identify which populations live closest
to construction and demolition landfills in North Carolina and therefore which 
populations stand to benefit the most from waste diversion in the construction 
industry. The focus on the construction industry was inspired by one of the 
group member's master's project which is focused on circularity and waste 
diversion in the construction industry. 


## 1.2 Research Question
Are construction and demolition landfills in North Carolina disproportionately  
more common in low income counties and/or counties with high minority  
populations?

## 1.3 Rational for Data 

####### what's the difference between this and section 2?

# 2 Dataset Information 

## 2.1 Source and Content of Data

########does this also count as table with dataset structure? Meghan askin John on 4/1

Dataset Names | Relevant Information | Source |
------------- | ---------------------| -----
Site Waste Facility  | Lists all landfills in NC and includes landfill type and address. | NC DEQ   https://deq.nc.gov/about/divisions/waste-management/solid-waste-section/solid-waste-facility-lists-presentations-and-annual-reports/solid-waste-facility-lists
CDC Social Vulnerability | Contains estimates of poverty and minority data by county. | Agency for Toxic Substances and Disease Registry https://www.atsdr.cdc.gov/placeandhealth/svi/documentation/SVI_documentation_2018.html
USA Counties Shape File | Spatial data frame that will establish the map of NC. | US Census   https://www.census.gov/geographies/mapping-files/time-series/geo/carto-boundary-file.html

## 2.2 Wrangling Process
SVI Variable Name | Variable Description | Units
-------------     | ---------------------| 
E_TOTPOP          | Population estimate,2014-2018 ACS | Count
E_POV             | Persons below poverty estimate,2014-2018 ACS | Count      
E_MINRTY          | Minority (all persons except white, non-Hispanic)estimate,2014-2018 ACS | Count

```{r include=FALSE, warning = FALSE}
#examine working directory
getwd()

#loading packages for spatial analysis 
library(tidyverse)
library(sf)
library(leaflet)
library(mapview)
library(RColorBrewer)
library(dplyr)
library(lubridate)

#loading packages for geocoding
#install.packages("tidygeocoder")
library(tidygeocoder)
library(tibble)
library(dplyr)

```


```{r data.loading.wrangling, warning=FALSE}
# read in landfill location raw data filter by waste type 'CD'
SW_FacilityList_CD <- read.csv("./Data/Raw/SW_FacilityList_actual.csv", stringsAsFactors = TRUE) %>%
  filter(Waste_Type == 'CD')

dim(SW_FacilityList_CD)

# read in landfill location data with correct facility coordinates using google maps
SW_Coordinates <- read.csv('./Data/Raw/CD_Facility_Coord.csv') %>%
  select(County:long)

dim(SW_Coordinates)

#read in social vulnerability raw data 
svi2018_nc_raw_full <- read.csv(
  'https://svi.cdc.gov/Documents/Data/2018_SVI_Data/CSV/States_Counties/NorthCarolina_COUNTY.csv')

#filter the 2018 SVI county-level dataset for County, FIPS, Location, E_TOTPOP, E_POV, and E_MINRTY
svi2018_nc_raw <- read.csv(
  'https://svi.cdc.gov/Documents/Data/2018_SVI_Data/CSV/States_Counties/NorthCarolina_COUNTY.csv',
  colClasses = c('FIPS' = 'factor')) %>% 
  select(COUNTY, FIPS, LOCATION, E_TOTPOP, E_POV, E_MINRTY)

#import tract-level svi data and change FIPS from character to factor 
svi2018_nc_tract_st <-st_read('./Data/Spatial/NorthCarolinaTract/SVI2018_NORTHCAROLINA_tract.shp')
svi2018_nc_tract_st$FIPS<-as.factor(svi2018_nc_tract_st$FIPS)
  
mapview(svi2018_nc_tract_st)

##check structure
str(svi2018_nc_raw)

#Read in USA counties shapefile and filter for NC - stateFP 37
counties_sf<- st_read('./Data/Spatial/cb_2018_us_county_20m.shp') %>% 
  filter(STATEFP == 37) 

```

geocode function extract addresses from the input dataframe and pass this data to the geo function which executes the geocoding queries. All extra arguments (...) given to geocode() are passed to geo() 
```{r geocoding}
#using site waste facility address to add latitude and longitude to sW_Facility_List_CD datafra
SW_FacilityList_CD_geo <- SW_FacilityList_CD %>%
  geocode(street = Address3, city = City, state = Sate, method="census")

dim(SW_FacilityList_CD_geo)

# remove rows with NAs in the longitude and latitude
SW_FacilityList_CD_complete <- na.omit(SW_FacilityList_CD_geo)

# merge geocoded site waste facilities with site waste facility data found using google maps
SW_FacilityList_complete <- rbind(SW_FacilityList_CD_complete, SW_Coordinates)

#permit numbers missing 3412-CDLF-1995 and 3505-TP-2020
dim(SW_FacilityList_complete)

#selecting columns we want to use for further analysis
SW_FacilityList_processed <- SW_FacilityList_complete %>% 
  select(County, Waste_Type, Permit_ID, Permit_Name, Address3, City, Zip, lat, long)

dim(SW_FacilityList_processed)

#writing csv to save wrangled data as a csv in processed folder 
write.csv(SW_FacilityList_processed, row.names = FALSE, file = "./Data/Processed/SW_FacilityList_processed.csv")

```


```{r joining.dataframes}
#Join the SVI attributes to the county spatial features
counties_sf_svi <- counties_sf %>% 
  left_join(svi2018_nc_raw, by = c("GEOID" = "FIPS"))

#Join the SVI_processed attributes to the county spatial features
counties_sf_svi_processed <- counties_sf %>% 
  left_join(svi2018_nc_processed, by = c("GEOID" = "FIPS"))

#Read our landfill points into a spatial dataframe, transform the datasets to use UTM 17 N, WGS84
SW_FacilityList_sf <- read_csv('./Data/Processed/SW_FacilityList_processed.csv') %>% 
  st_as_sf(coords = c('long','lat'), crs=32617)

```







