---
title: "EDA Final Report Spring 2022"
author: "Reino Hyyppa & Meghan Seyler"
date: "4/19/2022"
output:
  html_document:
    df_print: paged
---

Reminders for knitting commands in code chunks - can delete when we're done

* Knitting commands in code chunks:
 - `include = FALSE` - code is run, but neither code nor results appear in knitted file
 - `echo = FALSE`    - code not included in knitted file, but results are
 - `eval = FALSE`    - code is not run in the knitted file
 - `message = FALSE` - messages do not appear in knitted file
 - `warning = FALSE` - warnings do not appear...
 - `fig.cap = "..."` - adds a caption to graphical results

# 1. Introduction 

## 1.1 Context 

Waste diversion comes with global benefits including a decrease in  
global warming potential (GWP). However, waste diversion stands to especially 
benefit those populations living closest to landfills by decreasing local 
green house gas emissions and other hazards such as odor, smoke, noise, and  
water contamination.This project aims to identify which populations live closest
to construction and demolition landfills in North Carolina and therefore which 
populations stand to benefit the most from waste diversion in the construction 
industry. The focus on the construction industry was inspired by one of the 
group member's masters project which is focused on circularity and waste 
diversion in the construction industry. 


## 1.2 Research Questions
Question 1: Are construction and demolition landfills in North Carolina disproportionately  
located in counties with high minority and/or low income populations? 

Question 2: Are construction and demolition landfills in North Carolina disproportionately  
located in census tracts with high minority and/or low income populations?

Question 3: Is there significant differences in the distance to construction and demolition landfills among census tracts with high minority and/or low income populations? 

## 1.3 Rational for Data 
To investigate the research question we used three data sets: 

1. North Carolina Department of Environmental Quality Site Waste Facility data.  
This data set contains the addresses of each of the construction and 
demolition landfills in North Carolina. We chose this data set so that we could 
geocode the addresses into coordinates and map the landfills. 

2. CDC Social Vulnerability 2014-2018 data. This data set gave us the total  
population of each county and census tract as well as the number of minority   
residents and residents below the poverty line per county and per tract.  
Additionally, this data set was chosen because this data set has both location,  
via FIPS code, and SVI data it could be used to create a heat map of poverty and 
minority data by county. 

3. USA Counties Shape File from U.S. Census. This data set was chosen because it  
is the spatial dataframe needed to establish the map of North Carolina using the 
mapview function. 


# 2 Dataset Information 

## 2.1 Source and Content of Data

The source and general description of each data set are outlined in Table 1  
below. The variables of interest for each data set our outlined in Table 2 and 3.  
The variables in bold are variables that we created during the wrangling process.

**TABLE 1: Data set Description and Sources**

Dataset Names | Relevant Information | Source |
------------- | ---------------------| -----
Site Waste Facility  | Lists all landfills in NC and includes landfill type and address. | NC DEQ   https://deq.nc.gov/about/divisions/waste-management/solid-waste-section/solid-waste-facility-lists-presentations-and-annual-reports/solid-waste-facility-lists
CDC Social Vulnerability | Contains estimates of poverty and minority data by county. | Agency for Toxic Substances and Disease Registry https://www.atsdr.cdc.gov/placeandhealth/svi/documentation/SVI_documentation_2018.html
USA Counties Shape File | Spatial data frame that will establish the map of NC. | US Census   https://www.census.gov/geographies/mapping-files/time-series/geo/carto-boundary-file.html

**TABLE2: Relevant SVI Data Variables**
SVI Variable Name | Variable Description | Units
-------|--------------|-------|
E_TOTPOP  | Population estimate,2014-2018 ACS | Count
E_POV     | Persons below poverty estimate,2014-2018 ACS | Count      
E_MINRTY  | Minority estimate (all persons except white,non-Hispanic),2014-2018 ACS | Count
**POV_Percent**   | **(E_POV/E_TOTPOP)x100** | **Percentage**
**MINRTY_Percent**| **(E_MINRTY/E_TOTPOP)x100** | **Percentage**   

**TABLE3: Relevant Site Waste Facility Variables**
Landfill Location Variable Name | Variable Description 
-------|--------------|
Address3 | Landfill street name and area code 
City    | City name 
State     | State acronym  
**lat**       | **Landfill latitude** 
**long** | **Landfill longitude**

######### For processed data only: add table with dataset dimensions etc.This would be a table of column names and the data types, eg numeric, character,etc. Could be included with a list of mean values, missing data counts, etc

## 2.2 Wrangling Process

1. North Carolina Department of Environmental Quality Site Waste Facility data.  
    + Filter for only construction and demolition landfills 
    + use geocode function to turn addresses into latitude and longitude 
    + identify sites with N/As and inocorrect coordinates 
    + manually make an excel sheet with the pesky addresses and coordinates
    + merge the the original data set with the manually generate data set 
    
2. 

```{r include=FALSE, warning = FALSE}
#examine working directory
getwd()

#loading packages for spatial analysis 
library(tidyverse)
library(sf)
library(leaflet)
library(mapview)
library(RColorBrewer)
library(dplyr)
library(lubridate)

#loading packages for geocoding
#install.packages("tidygeocoder")
library(tidygeocoder)
library(tibble)
library(dplyr)

#loading packages for exploratory analysis 
library(knitr)
library(ggplot2)

```

```{r}
#import SVI tract-level data and change FIPS from character to factor 
svi2018_nc_tract_st <-st_read('./Data/Spatial/NorthCarolinaTract/SVI2018_NORTHCAROLINA_tract.shp')
svi2018_nc_tract_st$FIPS<-as.factor(svi2018_nc_tract_st$FIPS)
  
#mapview(svi2018_nc_tract_st)

#import and select columns of interest from SVI county data and change FIPS from character to factor
svi2018_nc_raw <- read.csv(
  'https://svi.cdc.gov/Documents/Data/2018_SVI_Data/CSV/States_Counties/NorthCarolina_COUNTY.csv',
  colClasses = c('FIPS' = 'factor')) %>% 
  select(COUNTY, FIPS, LOCATION, E_TOTPOP, E_POV, E_MINRTY)

```


# 3 Exploratory Analysis
(Rubric: flow between text and visualization is cohesive & relevant exploratory information is visualized)

Once the two data sets had been processed and converted into a spatial data frame, we conducted an exploratory analysis of our data visually using maps to display counties and census tracts for the state of North Carolina. We also created a map displaying the locations of construction and demolition landfills in North Carolina. Maps and visual analysis of the data was performed using 'ggplot' and the 'mapview' package.

```{r county map, fig.cap = "Map of North Carolina Counties", echo=FALSE}
counties_sf<- st_read('./Data/Spatial/cb_2018_us_county_20m.shp') %>% 
  filter(STATEFP == 37) 

mapview(counties_sf)

```

```{r tract map, fig.cap= "Map of North Carolina Census Tracts", echo=FALSE}

mapview(svi2018_nc_tract_st)

```


Additionally, we calculated the rate of minority and poverty in each county as a percent of total population to analyze the distribution of this data across North Carolina. We did this by dividing the total population by the number of low-income or minority households in each county. Then we created a chart of the top 10 counties ranked by rates of poverty and minority. We performed the same exploratory analysis on census tracts to provide a more localized and higher-resolution analysis of how this data is distributed across the state. Based on our calculations Robeson county, located in the southern part of the state, had the highest poverty and minority rates for North Carolina. 

```{r explore svi data, include=FALSE, echo=FALSE}
#check structure
#str(svi2018_nc_raw)

#check colnames
#colnames(svi2018_nc_raw)

#investigate what location column includes
#glimpse(svi2018_nc_raw$LOCATION)

#create percent poverty and minority values by county to explore which county's have the highest percentage minority and poverty populations 
povdividedbytotal_col<-(svi2018_nc_raw$E_POV/svi2018_nc_raw$E_TOTPOP)*100

minrtydividedbytotal_col<-(svi2018_nc_raw$E_MINRTY/svi2018_nc_raw$E_TOTPOP)*100

#add the percentage columns to the raw data
svi2018_nc_processed<-svi2018_nc_raw%>%
  add_column(Percent_POV = povdividedbytotal_col)%>%
  add_column(Percent_MINRTY = minrtydividedbytotal_col)

svi2018_nc_poverty_top10 <- svi2018_nc_processed %>%
    arrange(desc(Percent_POV))%>%
    slice(1:10)%>%
    relocate(Percent_POV, .before = FIPS)

view(svi2018_nc_poverty_top10)

svi2018_nc_minrty_top10 <- svi2018_nc_processed %>%
    arrange(desc(Percent_MINRTY))%>%
    slice(1:10)%>%
   relocate(Percent_MINRTY, .before = FIPS)

#create percent poverty and minority values by tract to explore which tracts have the highest minority and poverty populations by percentage 

povdividedbytotal_tract_col<-(svi2018_nc_tract_st$E_POV/svi2018_nc_tract_st$E_TOTPOP)*100

minrtydividedbytotal_tract_col<-(svi2018_nc_tract_st$E_MINRTY/svi2018_nc_tract_st$E_TOTPOP)*100

svi2018_nc_tract_st<-svi2018_nc_tract_st%>%
  add_column(Percent_POV = povdividedbytotal_tract_col)%>%
  add_column(Percent_MINRTY = minrtydividedbytotal_tract_col)

svi2018_nc_tract_st<-svi2018_nc_tract_st%>%
  select(COUNTY, FIPS, LOCATION, E_TOTPOP, E_POV, E_MINRTY,Percent_POV,Percent_MINRTY)

svi2018_nc_tract_st_top10 <- svi2018_nc_tract_st %>%
    arrange(desc(Percent_POV))%>%
    slice(1:10)%>%
    relocate(Percent_POV, .before = FIPS)

svi2018_nc_tract_st_top10 <- svi2018_nc_tract_st %>%
    arrange(desc(Percent_MINRTY))%>%
    slice(1:10)%>%
   relocate(Percent_MINRTY, .before = FIPS)

```


```{r table 1, fig.cap= "Table of Top 10 Counties Ranked by Percent Poverty"}
knitr::kable(head(svi2018_nc_poverty_top10[1:10,1:2]), 
             "simple", 
             col.names = c("County","Poverty Percentage"),
             align = 'c',
             caption = "Counties by Highest Poverty Percentage in North Carolina"
)
```

```{r table 2,  fig.cap= "Table of Top 10 Counties Ranked by Percent Minority"}
knitr::kable(head(svi2018_nc_minrty_top10[1:10,1:2]), 
             "simple", 
             col.names = c("County","Minority Percentage"),
             align = 'c',
             caption = "Counties by Highest Percentage of Minority Populations in North Carolina"
) 
```

We also created two separate lollipop plots using the 'ggplot' package to visualize and explore the relationship between counties and percentages of both poverty and low-income populations for North Carolina. 


```{r exploration plots, fig.cap = "Pecentage of Poverty by County"}
poverty_plot<-svi2018_nc_processed%>%
  arrange(Percent_POV)%>%
  mutate(COUNTY=factor(COUNTY, COUNTY))%>%
    ggplot(aes(x=COUNTY, y=Percent_POV)) +
      geom_point(color="lightgreen") + 
      geom_segment( aes(x=COUNTY, xend=COUNTY, y=0, yend=Percent_POV))+
      coord_flip()+
      ggtitle("Percentage of Poverty by County in North Carolina") +
      xlab("County Name")+
      ylab("Poverty Population Percentage")+
      theme_minimal()+
      theme(
        axis.text = element_text(size = 5, face='bold', angle = 30, hjust = 1, vjust =0.5) ,
        panel.grid.major.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.y = element_blank(),
  )

poverty_plot
```

```{r minority lolliplot plot, fig.cap = "Percent of Minority by County"}
minority_plot<-svi2018_nc_processed%>%
  arrange(Percent_MINRTY)%>%
  mutate(COUNTY=factor(COUNTY, COUNTY))%>%
    ggplot(aes(x=COUNTY, y=Percent_MINRTY)) +
      geom_point(color="69b3a2") + 
      geom_segment( aes(x=COUNTY, xend=COUNTY, y=0, yend=Percent_MINRTY))+
      coord_flip()+
      ggtitle("Percentage of Minority Populations by County in North Carolina") +
      xlab("County Name")+
      ylab("Minority Population Percentage")+
      theme_minimal()+
      theme(
        axis.text = element_text(size = 5, face='bold'),
        panel.grid.major.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.y = element_blank(),
        plot.subtitle = element_text("By Reino Hyyppa and Meghan Seyler"),
  )
minority_plot
```

Finally, we wanted to determine the nearest construction and demolition landfill to each census tracts and calculate the distances between each point. First, using the st_centroid function in the 'sf' package in r we were able to locate the centroid of each county tract. Next, we used the st_nn function in the 'nngeo' package to identify the nearest landfill to each centroid and calculate the distance between each point using the euclidean distance method. In the next section, we will use these distance measurements to explore the relationship between construction and demolition landfills and percentages of minority and low income populations for each census tract in North Carolina. 

# 4 Analysis

## 4.1 Q1: Are construction and demolition landfills in North Carolina disproportionately located in counties with high minority and/or low income populations? 

First, we conducted a visual analysis to understand the relationship between construction and demolition landfills and counties with high rates of low-income populations in North Carolina. Using the 'mapview' package we created a heatmap to represent low-income populations graphically (Figure ). Counties with high rates of low-income populations are shaded blue, while counties with low rates of low-income populations are colored red. Next, we joined the heat map of percetage for each county with the map of construction and demolition landfills in North Carolina to visualize their relationship. 

```{r county level, fig.cap = "Heat Map of Percentage Poverty by Counties and C&D Landfill Sites"}

#View minortiy and poverty counts with mapview
#mapview(counties_sf_svi_processed, 
        #zcol = 'E_POV', 
       # col.regions = brewer.pal(2, 'RdBu')) + 
  #mapview(landfill_sites_sf, cex = 3, col.regions="green")

#mapview(counties_sf_svi_processed, 
    #    zcol = 'E_MINRTY', 
     #   col.regions = brewer.pal(2, 'RdBu')) + 
 # mapview(landfill_sites_sf, cex = 3, col.regions="green")

#view minority and poverty percentages with mapview
mapview(counties_sf_svi_processed, 
        zcol = 'Percent_POV', 
        col.regions = brewer.pal(2, 'RdBu')) + 
  mapview(landfill_sites_sf, cex = 3, col.regions="green")
```

Additionally, we performed a visual analysis to examine the relationship between construction and demolition landfills and counties with high rates of minority populations using a heat map (Figure ). Counties with high rates of minority populations are colored blue, and counties with low rates of minority populations are shaded red. 

```{r poverty by county, fig.cap = "Heat Map of Percentage Minority by Counties and C&D Landfill Sites"}
mapview(counties_sf_svi_processed, 
        zcol = 'Percent_MINRTY', 
        col.regions = brewer.pal(2, 'RdBu')) + 
  mapview(landfill_sites_sf, cex = 3, col.regions="green")

```

# Q2: Are construction and demolition landfills in North Carolina disproportionately located in census tracts with high minority and/or low income populations?

The next step of our analysis, was exploring the relationship between construction and demolition landfills and social vulnerability data at the census tract level for North Carolina. Census tracts are a subdivision of counties and are comparable in size to one another, and therefore provide a more localized measure of the rates of low-income and minority populations. Using a similar approach to  our analysis of the relationship between construction and demolition landfills and counties with high rates of low-income and minority populations, we created heat maps of social vulnerability data for each census tract. 

First, we created a map of the percentage of low-income population for each census tract using the 'mapview' package. We joined this map with the map of construction and demolition landfills to visualize the their relationship (Figure ). Census tracts with high rates of low-income population are colored and blue and census tracts with low rates of low-income population are shaded red.

```{r poverty by census tract, fig.cap = "Heat Map of Percentage Poverty by Census Tract and C&D Landfill Sites"}
#view minority and poverty percentages with mapview
mapview(svi2018_nc_tract_st, 
        zcol = 'Percent_POV', 
        col.regions = brewer.pal(2, 'RdBu')) + 
  mapview(landfill_sites_sf, cex = 3, col.regions="green", alpha = 1)

```

Next, we performed the same analysis to determine the relationship between construction and demolition landfills and census tracts with high rates of minority using a heat map. Again, census tracts with high rates of minority population are colored blue, and census tracts with low-minority percentages are shaded red. 
```{r minority by census tract, fig.cap ="Heat Map of Percentage Minorityy by Census Tract and C&D Landfill Sites }
mapview(svi2018_nc_tract_st, 
        zcol = 'Percent_MINRTY', 
        col.regions = brewer.pal(2, 'RdBu')) + 
  mapview(landfill_sites_sf, cex = 3, col.regions="green")
```

# Q3: Is there significant differences in the distance to construction and demolition landfills among census tracts with high minority and/or low income populations? 

In addition, we wanted to determine if there is a significant relationship between social vulnerability data for each census tract and the proximity to the nearest landfill. To do this we created both a linear model to examine the relationship between these variables, as well as a Kendall Non-parametric Correlation test. 

First, we conducted a correlation test using the cor.test function in the 'stats' package in r to test the relationship between social vulnerability data for each census tract and their distances to the nearest landfill. We used the Kendall Correlation test for non-parametric data since the the measure of social vulnerability was ordinal data using percentages and does not follow a normal distribution. We performed the Kendall Correlation test on both poverty and minority data to determine the relationship between census tracts and their proximity to the nearest landfill.

Finally, we analyzed the relationship between census tracts and their proximity to landfills using the lm function in the 'stats' package. We used the distance between each census tract centroid and the nearest landfill as the dependent variable and social vulnerability characteristics (percentage of poverty and low-income population for each census tract) as the independent variables. 

```{r linear model, echo=FALSE, include=FALSE}

#Convert to UTM Zone 17 (crs = 26917)
svi2018_tract_utm <- st_transform(svi2018_nc_tract_st, crs = 26917)
landfill_utm <- st_transform(landfill_sites_sf, crs = 26917)

# identify census tract centroids using st_centroid function
tract_centroids <- st_centroid(svi2018_tract_utm)
mapview(tract_centroids) 

# install packages for nearest neighbor calculation
#install.packages('nngeo')
library(nngeo)

# identify nearest landfill to each tract centroid and calculate distance
landfill_nn <- st_nn(x=tract_centroids, y=landfill_utm, returnDist = TRUE, k = 1) %>% 
  set_names("Site", "Distance") %>% 
  map_df(., unlist)

# combine centroid-landfill distances with SVI data set 
svi2018_nc_tract_st <- cbind(svi2018_nc_tract_st, landfill_nn)

# linear model for low-income population and distance to landfill
pov_regression <- lm(data = svi2018_nc_tract_st, Distance ~ Percent_POV)
summary(pov_regression)


# linear model for minority populations and distance to landfill 
min_regression <- lm(data = svi2018_nc_tract_st, Distance ~ Percent_MINRTY)
summary(min_regression)

# check diagnostic plots
par(mfrow = c(1,1))
plot(min_regression)

# remove outlier (interestingly all the outliers are from the same county)
inx.o <- c(342, 341, 340, 339, 338, 337)
m2_out <- with(svi2018_nc_tract_st[-inx.o, ], lm(Distance~Percent_MINRTY))

# removing outliers did not meet the assumptions of the model
par(mfrow=c(1,1))
plot(m2_out)

# log transform social vulnerability data
svi2018_nc_tract_st <- 
  mutate(svi2018_nc_tract_st, log_min = log(svi2018_nc_tract_st$Percent_MINRTY),
         svi2018_nc_tract_st, log_pov = log(svi2018_nc_tract_st$Percent_POV))

# run linear model on log transformed minority data
logMin_regression <- lm(data = svi2018_nc_tract_st, Distance ~ log_min)
summary(min_regression)

# try to remove any NAs
svi2018_nc_tract_st <- na.omit(svi2018_nc_tract_st) 

# rerun the model

logMin_regression2 <- lm(data = svi2018_nc_tract_st, Distance ~ log_min)
summary(logMin_regression2)

# Test for correlation using the Kendall Rank correlation test
cor.test(svi2018_nc_tract_st$Percent_MINRTY, svi2018_nc_tract_st$Distance, method = "kendall")
cor.test(svi2018_nc_tract_st$Percent_POV, svi2018_nc_tract_st$Distance, method = "kendall")
```

# 4 Visualization

To visualize the relationship between both percentages of minority and low-income populations and their proximity to the nearest landfill for each census tract we create a bubble plot using the 'ggplot' package (Figure ). In the figure below, we graphed the percentage low-income population for each census tract on the y-axis and the percentage minority population on the y-axis. Each point on the plot is sized by the distance from each census tract to the nearest landfill.  

```{r plot data, fig.cap = "Percentage of Low-income vs. Minority Population for North Carolina}
ggplot(svi2018_nc_tract_st, aes(Percent_MINRTY, Percent_POV, size = Distance)) +
  geom_point(alpha = 0.5)
```

# 5 Results

# 5.1 Q1: Are construction and demolition landfills in North Carolina disproportionately located in counties with high minority and/or low income populations? 

# 5.2  Q2: Are construction and demolition landfills in North Carolina disproportionately located in census tracts with high minority and/or low income populations?

# 5.3 Q3: Is there significant differences in the distance to construction and demolition landfills among census tracts with high minority and/or low income populations? 

There is evidence that the proximity to the nearest construction and demolition landfill is correlated with the percentage of low-income and minority populations for each census tract. In particular, the Kendall Correlation test revealed that there is a significant inverse relationship between proximity to the nearest landfill and the percentage of minority population for each census tract (p-value < 2.2e-16, tau = -0.13). In addition, there is a significant positive correlation between the proximity to the nearest landfill and the percentage of low-income population in each census tract (p-value = 0.001, tau = 0.04649479). 

There is also a significant relationship between the proximity to the nearest landfill and the percentage of minority population for each census tract (r^2 = 0.03, p-value < 2.2e-16). The coefficient for percentage minority population is -125.57, which in context means that for every 1% increase in minority population, the distance from the census tract centroid to the nearest landfill decreases by 125.57 meters. There was not a significant relationship between landfill proximity ad percentage of low-income population for North Carolina (p-value = 0.1921).  

# 6 Summary and Conclusion





