---
title: "SeylerHyyppa_Final Project"
author: "Reino Hyyppa & Meghan Seyler"
date: "3/31/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Reminders for knitting commands in code chunks - can delete when we're done

* Knitting commands in code chunks:
 - `include = FALSE` - code is run, but neither code nor results appear in knitted file
 - `echo = FALSE`    - code not included in knitted file, but results are
 - `eval = FALSE`    - code is not run in the knitted file
 - `message = FALSE` - messages do not appear in knitted file
 - `warning = FALSE` - warnings do not appear...
 - `fig.cap = "..."` - adds a caption to graphical results

```{r include=FALSE}
#examine working directory
getwd()

#loading packages for spatial analysis 
library(tidyverse)
library(sf)
library(leaflet)
library(mapview)
library(RColorBrewer)

```


### 2.2.1 Read in and explore NC counties

Below we read in the USA counties shapefile, filtering for just the NC features (NC has a state FIPS code of "37"...). We also see that `sf` plays nice with "tidyverse" syntax (e.g. pipes) and functions (e.g. filter). The `sf` package also includes some new spatial methods for exploring our data (e.g. `st_bbox` which draws a bounding box around all spatial features).

```{r read.filtered.records.from.shapefile.to.sf.object}
#
counties_sf<- st_read('./Data/Spatial/cb_2018_us_county_20m.shp') %>% 
  filter(STATEFP == 37) #Filter for just NC Counties

#
colnames(counties_sf)

#
st_crs(counties_sf) 


#
nrow(counties_sf)

#Reveal the extent of this dataset via the st_bbox() function
st_bbox(counties_sf)

#View the data
head(counties_sf)

#Plot the data, colored by area of land in each county
mapView(counties_sf, zcol = "AWATER")









